---
title: "Analysis of methylation on spike-ins - CAPS data"
author: "Zhiyuan"
date: "3/9/2020 (last modified: `r Sys.Date()`)"
output: 
  prettydoc::html_pretty:
    theme: architect
    highlight: github
# output: 
#   pdf_document:
#     latex_engine: xelatex
#     df_print: paged
#     number_sections: yes
#     toc: yes
# fontsize: "11pt"
# keep_tex: False
# header-includes:
#   - \usepackage{fontspec}
#   - \setmainfont{Arial}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(ggplot2)
```

```{r read_mods}
mods_h <- read.delim("../astair_output/spike_in_new/YL712508hm_sorted_spikein_clipOverlap_mCtoT_all.mods",as.is = T)
mods_m <- read.delim( "../astair_output/spike_in_new/YL711505m_sorted_spikein_clipOverlap_mCtoT_all.mods",as.is = T)
mods_f <- read.delim( "../astair_output/spike_in_new/YL708508f_sorted_spikein_clipOverlap_mCtoT_all.mods",as.is = T)
mods_c <- read.delim( "../astair_output/spike_in_new/YL709502c_sorted_spikein_clipOverlap_mCtoT_all.mods",as.is = T)

mods_list <- list(mods_m,  # tapsbeta
                  mods_h,  # caps
                  mods_f,  # ps
                  mods_c)  # psc

methods <- c("TAPSbeta","CAPS", "PS", "PSc")

for(i in 1:4){
  mods_list[[i]] <- mods_list[[i]][(mods_list[[i]]$MOD + mods_list[[i]]$UNMOD) > 0,]
  colnames(mods_list[[i]])[1] <- "CHROM"
  mods_list[[i]]$MOD_LEVEL <- as.numeric(mods_list[[i]]$MOD_LEVEL)
}
```

```{r 2kb_spikein}
spikein <- "2kb_2"
tmp_sum <- data.frame(methods = methods,
                      conversion = NA)
for(i in 1:4){
  mods <- mods_list[[i]]
  tmp_mods <- mods[mods$CHROM == spikein & mods$CONTEXT == "CpG",]
  tmp_sum$conversion[i] <- mean(tmp_mods$MOD_LEVEL[(tmp_mods$MOD + tmp_mods$UNMOD) >= 3])
    # sum(tmp_mods$MOD)/(sum(tmp_mods$MOD) +  sum(tmp_mods$UNMOD))
}
df_2kb2 <- tmp_sum
knitr::kable(df_2kb2, digits = 4, caption = "Summary of 2kb_2")
```

```{r lambda}
spikein <- "J02459.1"
tmp_sum <- data.frame(methods = methods,
                      conversion = NA)
for(i in 1:4){
  mods <- mods_list[[i]]
  tmp_mods <- mods[mods$CHROM == spikein & mods$CONTEXT == "CpG",]
  # tmp_sum$conversion1[i] <- mean(tmp_mods$MOD_LEVEL) ## methods 1 calculate average
  # tmp_sum$conversion[i] <- mean(tmp_mods$MOD_LEVEL[(tmp_mods$MOD + tmp_mods$UNMOD) >= 3])
  tmp_sum$conversion[i] <- mean(tmp_mods$MOD_LEVEL[(tmp_mods$MOD + tmp_mods$UNMOD) >= 3])
    # sum(tmp_mods$MOD)/( sum(tmp_mods$MOD) +  sum(tmp_mods$UNMOD))
}

df_lambda <- tmp_sum
knitr::kable(df_lambda, digits = 5, caption = "Summary of lambda DNA")
```

```{r 144fc}
spikein <- "144fc"
tmp_sum <- data.frame(methods = methods,
                      conversion_CpG = NA,
                      conversion_CAA = NA)
loc <- c(87, 92, 100, 110) # , 117

for(i in 1:4){
  mods <- mods_list[[i]]
  tmp_mods <- mods[mods$CHROM == spikein & mods$END %in% loc,]
  tmp_sum$conversion_CpG[i] <- mean(tmp_mods$MOD_LEVEL[(tmp_mods$MOD + tmp_mods$UNMOD) >= 3])
    # sum(tmp_mods$MOD)/(sum(tmp_mods$MOD) +  sum(tmp_mods$UNMOD))
}

loc <- 117
for(i in 1:4){
  mods <- mods_list[[i]]
  tmp_mods <- mods[mods$CHROM == spikein & mods$END %in% loc, , drop=FALSE]
  tmp_sum$conversion_CAA[i] <- (tmp_mods$MOD)/((tmp_mods$MOD) + (tmp_mods$UNMOD))
}

df_144fc <- tmp_sum
knitr::kable(df_144fc, digits = 4, caption = "Summary of 144fc spike in containing four modified CpGs and one modified CAA.")
```

```{r N5mCNN, eval=FALSE}
spikein <- "synthetic_N5mCNN"

tmp_sum <- data.frame(methods = methods,
                      conversion_site1 = NA,
                      conversion_site2 = NA)
loc <- 51
for(i in 1:4){
  mods <- mods_list[[i]]
  tmp_mods <- mods[mods$CHROM == spikein & mods_m$END %in% loc,]
  tmp_sum$conversion_site1[i] <- tmp_mods$MOD_LEVEL
}

loc <- 111
for(i in 1:4){
  mods <- mods_list[[i]]
  tmp_mods <- mods[mods$CHROM == spikein & mods_m$END %in% loc,]
  tmp_sum$conversion_site2[i] <- tmp_mods$MOD_LEVEL
}

df_N5mCNN <- tmp_sum
knitr::kable(df_N5mCNN, digits = 6, caption = "Summary of N5mCNN containing one 5mC and one 5hmC")
```

```{r unmodified}
spikein <- "unmodified_2kb"

tmp_sum <- data.frame(methods = rep(methods, each = 3),
                      context = rep(c("CpG","CHG","CHH"), 4),
                      conversion = NA)

for(i in 1:4){
  for(j in c("CpG","CHG","CHH")) {
    tmp_mods <- mods_list[[i]][mods_list[[i]]$CHROM == spikein & mods_list[[i]]$CONTEXT == j,]
    tmp_sum$conversion[tmp_sum$methods == methods[i] & tmp_sum$context == j] <- mean(tmp_mods$MOD_LEVEL[(tmp_mods$MOD+tmp_mods$UNMOD)>=3])
  }
}

df_unmodified <- tmp_sum
knitr::kable(df_unmodified, digits = 5, caption = "Summary of unmodified spike in")
```

```{r}
df_unmodified <- df_unmodified[df_unmodified$context == "CpG", c("methods","conversion")]
write.table(df_unmodified, "false_positive_table.txt", sep = "\t", quote = F, row.names = F, col.names = T)

```


```{r unmodified_CpG_only}
spikein <- "unmodified_2kb"

tmp_sum <- data.frame(methods = rep(methods, each = 1),
                      conversion = NA)

for(i in 1:4){
    tmp_mods <- mods_list[[i]][mods_list[[i]]$CHROM == spikein & mods_list[[i]]$CONTEXT == "CpG",]
    tmp_sum$conversion[i] <- mean(tmp_mods$MOD_LEVEL[(tmp_mods$MOD + tmp_mods$UNMOD) >= 3])
      # sum(tmp_mods$MOD)/(sum(tmp_mods$MOD) +  sum(tmp_mods$UNMOD))
}

df_unmodified <- tmp_sum
knitr::kable(df_unmodified, digits = 5, caption = "Summary of unmodified spike in")
```


# Conversion rate, false positive rate and mapping rate


```{r N5mCNN results}
N5mCNN_res <- read.csv("../../results/2020-07-26_spikein/synthetic_N5mCNN_results.csv", as.is = T)
```


## TAPS-beta

N5mCNN: Top strand 0.948047631

5mC: lambda 0.9676 (use this one; CpG)

5hmC: N5mCNN bottom strand (CpG only)

Unmodified: unmodified_2kb (both CpG and CH in one column)

```{r tapsbeta_conversion_rate}
df_plot <- data.frame(type = c("5mC","5hmC"),
                      cr = c(df_lambda$conversion[df_lambda$methods == "TAPSbeta"],
                             N5mCNN_res$Statistic[N5mCNN_res$Methods == "TAPSbeta" & 
                                                    N5mCNN_res$Strand == "Bottom_strand"&
                                                    N5mCNN_res$Value == "ratio_modified_CG"]))
df_plot$type <- factor(df_plot$type, levels = c("5mC","5hmC"))

ggplot(df_plot, aes(x = type, y = cr*100)) + geom_bar(stat = "identity", fill = "grey", col = "grey20") + 
    geom_text(aes(label=round(cr * 100, 1)), vjust=-0.4) +
  theme_light() + xlab("") + ylab("Fraction converted cytosines on CpGs (%)") + theme(axis.text = element_text(color = "black"))+ ylim(0,100)

ggsave("plots/basic_statistics20200726/TAPSbeta_conversion.pdf",width = 2, height = 4)
```


```{r tapsbeta_Unmodified_rate}
df_plot <- data.frame(type = c("Unmodified"),
                      cr = c(df_unmodified$conversion[df_unmodified$methods == "TAPSbeta"]))

ggplot(df_plot, aes(x = type, y = cr*100)) + geom_bar(stat = "identity", fill = "grey40", col = "grey20") + 
    geom_text(aes(label=round(cr * 100, 2)), vjust=-0.4) +
  theme_light() + xlab("") + ylab("False positive rate (%)") + theme(axis.text = element_text(color = "black")) +
  ylim(c(0,1))

ggsave("plots/basic_statistics20200726/TAPSbeta_unmofified_conversion.pdf",width = 1.4, height = 4)
```

## CAPS

5mC: lambda (CpG)

5hmC: N5mCNN bottom strand (CpG only)

Unmodified: unmodified_2kb

```{r caps_conversion_rate}
df_plot <- data.frame(type = c("5mC","5hmC"),
                      cr = c(df_lambda$conversion[df_lambda$methods == "CAPS"],
                             N5mCNN_res$Statistic[N5mCNN_res$Methods == "CAPS" & 
                                                    N5mCNN_res$Strand == "Bottom_strand"&
                                                    N5mCNN_res$Value == "ratio_modified_CG"]))
df_plot$type <- factor(df_plot$type, levels = c("5mC","5hmC"))

ggplot(df_plot, aes(x = type, y = cr*100)) + geom_bar(stat = "identity", fill = "grey", col = "grey20") + 
    geom_text(aes(label=round(cr * 100, 2)), vjust=-0.4) +
  theme_light() + xlab("") + ylab("Fraction converted cytosines on CpGs (%)") + theme(axis.text = element_text(color = "black")) + ylim(0,100)

ggsave("plots/basic_statistics20200726/CAPS_conversion.pdf",width = 2, height = 4)
```


```{r caps_Unmodified_rate}
df_plot <- data.frame(type = c("Unmodified"),
                      cr = c(df_unmodified$conversion[df_unmodified$methods == "CAPS"]))

ggplot(df_plot, aes(x = type, y = cr*100)) + geom_bar(stat = "identity", fill = "grey40", col = "grey20") + 
    geom_text(aes(label=round(cr * 100, 2)), vjust=-0.4) +
  theme_light() + xlab("") + ylab("False positive rate (%)") + theme(axis.text = element_text(color = "black")) +
  ylim(c(0,1))

ggsave("plots/basic_statistics20200726/CAPS_unmofified_conversion.pdf",width = 1.4, height = 4)
```

## PS 

5mC: lambda (CpG)

5hmC: N5mCNN bottom strand (CpG only)

fC: 144fc

caC: 2kb_2

Unmodified: unmodified_2kb


```{r PS_conversion_rate}
df_plot <- data.frame(type = c("5mC","5hmC","5fC","5caC"),
                      cr = c(df_lambda$conversion[df_lambda$methods == "PS"],
                             N5mCNN_res$Statistic[N5mCNN_res$Methods == "PS" & 
                                                    N5mCNN_res$Strand == "Bottom_strand"&
                                                    N5mCNN_res$Value == "ratio_modified_CG"],
                             df_144fc$conversion_CpG[df_144fc$methods == "PS"],
                             df_2kb2$conversion[df_2kb2$methods == "PS"]))
df_plot$type <- factor(df_plot$type, levels = c("5mC","5hmC","5fC","5caC"))

ggplot(df_plot, aes(x = type, y = cr*100)) + geom_bar(stat = "identity", fill = "grey", col = "grey20") + 
    geom_text(aes(label=round(cr * 100, 2)), vjust=-0.4) +
  theme_light() + xlab("") + ylab("Fraction converted cytosines on CpGs (%)") + theme(axis.text = element_text(color = "black")) + ylim(0,100)

ggsave("plots/basic_statistics20200726/PS_conversion.pdf",width = 3.8, height = 4)
```


```{r PS_Unmodified_rate}
df_plot <- data.frame(type = c("Unmodified"),
                      cr = c(df_unmodified$conversion[df_unmodified$methods == "PS"]))

ggplot(df_plot, aes(x = type, y = cr*100)) + geom_bar(stat = "identity", fill = "grey40", col = "grey20") + 
    geom_text(aes(label=round(cr * 100, 2)), vjust=-0.4) +
  theme_light() + xlab("") + ylab("False positive rate (%)") + theme(axis.text = element_text(color = "black")) +
  ylim(c(0,1))

ggsave("plots/basic_statistics20200726/PS_unmofified_conversion.pdf",width = 1.4, height = 4)
```


## PSc

5mC: lambda (CpG)

5hmC: N5mCNN bottom strand (CpG only)

fC: 144fc

caC: 2kb_2

Unmodified: unmodified_2kb

```{r PSc_conversion_rate}
df_plot <- data.frame(type = c("5mC","5hmC","5fC","5caC"),
                      cr = c(df_lambda$conversion[df_lambda$methods == "PSc"],
                             N5mCNN_res$Statistic[N5mCNN_res$Methods == "PSc" & 
                                                    N5mCNN_res$Strand == "Bottom_strand"&
                                                    N5mCNN_res$Value == "ratio_modified_CG"],
                             df_144fc$conversion_CpG[df_144fc$methods == "PSc"],
                             df_2kb2$conversion[df_2kb2$methods == "PSc"]))
df_plot$type <- factor(df_plot$type, levels = c("5mC","5hmC","5fC","5caC"))

ggplot(df_plot, aes(x = type, y = cr*100)) + geom_bar(stat = "identity", fill = "grey", col = "grey20") + 
    geom_text(aes(label=round(cr * 100, 3)), vjust=-0.4) +
  theme_light() + xlab("") + ylab("Fraction converted cytosines on CpGs (%)") + theme(axis.text = element_text(color = "black")) + ylim(0,100)

ggsave("plots/basic_statistics20200726/PSc_conversion.pdf",width = 3.8, height = 4)
```


```{r PSc_Unmodified_rate}
df_plot <- data.frame(type = c("Unmodified"),
                      cr = c(df_unmodified$conversion[df_unmodified$methods == "PSc"]))

ggplot(df_plot, aes(x = type, y = cr*100)) + geom_bar(stat = "identity", fill = "grey40", col = "grey20") + 
    geom_text(aes(label=round(cr * 100, 3)), vjust=-0.4) +
  theme_light() + xlab("") + ylab("False positive rate (%)") + theme(axis.text = element_text(color = "black")) +
  ylim(c(0,1))

ggsave("plots/basic_statistics20200726/PSc_unmofified_conversion.pdf",width = 1.4, height = 4)
```


 
